---
title: "09 - Antigen specificity (database analysis)"
authors: "Maria Chernigovskaya and Teresa Rubio"
date: "March, 2021"
output:
    html_document:
      code_folding: hide
params:
  inputDir: !r getwd()
  workDir: !r getwd()
  outputDir: !r getwd()
  sampleInfo: !r NULL
  chain: !r NULL
  specie: !r NULL
---

# Packages
```{r, warning=F, message=FALSE}
library(immunarch)
library(reshape2)
library(openxlsx)
```

# Work enviroment
```{r}
# Define output directory and TCR chain
sampleInfoFile <- params$sampleInfo
dir.create(params$outputDir, recursive=T)
chain <- params$chain

# Load sample info
SampleInfo <- read.table(sampleInfoFile, sep = ",", header = T, check.names = F)
levels <- unique(SampleInfo$Group)
SampleInfo$Group <- factor(SampleInfo$Group, levels = levels)
```

# McPAS
The database was last updated on: March 6, 2021. #TRA+TRB ddbb:August 5, 2021.
Downloaded on: March 12, 2021. #January 21, 2022
```{r}
mcpas_db <- read.csv("McPAS-TCR.csv", stringsAsFactor = FALSE, sep = ",")

mcpas_db <- mcpas_db[!is.na(mcpas_db$CDR3.beta.aa),]
# Correcting database typos
mcpas_db$Pathology <- gsub("IBD\\).", "IBD)", mcpas_db$Pathology)
mcpas_db$Pathology <- gsub("M.Tuberculosis", "M. tuberculosis", mcpas_db$Pathology)
mcpas_db$Pathology <- gsub("Hepatitis C virus \\(HCV\\)", "Hepatitis C virus", mcpas_db$Pathology)
mcpas_db$Pathology <- gsub("HTLV-1 \\(Chronic", "HTLV-1", mcpas_db$Pathology)
mcpas_db$Pathology <- gsub("Psoriatic arthritis", "Psoriatic Arthritis", mcpas_db$Pathology)
# Filtering database
mcpas_db_specie <- subset(mcpas_db, Species == params$specie & nchar(CDR3.beta.aa) > 4)
# mcpas_db_specie <- subset(mcpas_db_specie, Category == "Autoimmune" | Category == "Pathogens")
mcpas_db_specie <- subset(mcpas_db_specie, !is.na(Pathology))
```

# VDJdb
Last updated on 02 February, 2021. #TRA+TRB ddbb: 05 September, 2021
Downloaded on: March 12, 2021. #January 21, 2022
```{r}
VDJdb <- read.delim("VDJdb-TCR.tsv")

if (params$specie == "Human") {
  specie <- "HomoSapiens"
} else {
  specie <- "MusMusculus"
}

vdjdb_specie <- subset(VDJdb, Species == specie & nchar(CDR3) > 4)
vdjdb_specie <- subset(vdjdb_specie, Gene == chain)
vdjdb_specie <- subset(vdjdb_specie, !is.na(Epitope.species))
```

# CDR3 sets by experimental group
```{r}
cdrsets <- list()
immdata <- immunarch::repLoad(file.path(params$inputDir, paste0("clones_",chain,"_filtered")))

# # control group
# control <- immdata$data[c(SampleInfo$SampleID[SampleInfo$Group == "control"])]
# control_cdrs <- unique(do.call(rbind, lapply(control, function(x) x[,"CDR3.aa",drop=F])))
# cdrsets$control <- control_cdrs$CDR3.aa
# 
# # withoutMHE group
# withoutMHE <- immdata$data[c(SampleInfo$SampleID[SampleInfo$Group == "withoutMHE"])]
# withoutMHE_cdrs <- unique(do.call(rbind, lapply(withoutMHE, function(x) x[,"CDR3.aa",drop=F])))
# cdrsets$withoutMHE <- withoutMHE_cdrs$CDR3.aa
# 
# # withMHE group
# withMHE <- immdata$data[c(SampleInfo$SampleID[SampleInfo$Group == "withMHE"])]
# withMHE_cdrs <- unique(do.call(rbind, lapply(withMHE, function(x) x[,"CDR3.aa",drop=F])))
# cdrsets$withMHE <- withMHE_cdrs$CDR3.aa
save.image(file = "my_work_space.RData")
cdrsets <- setNames(lapply(levels, function(i){
              group <- immdata$data[c(SampleInfo$SampleID[SampleInfo$Group == i])]
              group_cdrs <- unique(do.call(rbind, lapply(group, function(x) x[,"CDR3.aa",drop=F])))
              group_cdrs$CDR3.aa
            }), nm = levels)
```


# Statistical test
## ref db = mcpas
The p-values are adjusted for multiple testing using Benjamini-Hochberg FDR correction considering both the number of diseases and the number of sample groups tested.
```{r}
# db : reference database (mcpas)
# cdr3 : data (CDR3 sets)
# disease : disease name (e.g. HIV, CMV, etc)
get_stats <- function(db, cdr3, disease) {
  disease_cdr3 <- unique(db$CDR3.beta.aa[db$Pathology == disease])
  cont_table <- table(unique(db$CDR3.beta.aa) %in% cdr3, unique(db$CDR3.beta.aa) %in% disease_cdr3)
  rownames(cont_table) <- c("not_in_data", "in_data")
  colnames(cont_table) <- c("not_in_disease","in_disease")
  pval <- fisher.test(cont_table, alternative = "great")$p.value
  return(pval)
}

pvalues <- do.call(cbind, lapply(cdrsets, function(cdr3set){
  diseases <- unique(mcpas_db_specie$Pathology)
  pvals <- rep(NA, length(diseases))

  for (i in 1:length(diseases)) {
    pvals[i] <- get_stats(db = mcpas_db_specie, cdr3 = cdr3set, disease = diseases[i])
    names(pvals)[i] <- diseases[i]
  }

  return(pvals)

}))

adj_pval <- matrix(p.adjust(pvalues, method = "fdr"), nrow = nrow(pvalues), ncol = length(levels))
rownames(adj_pval) <- rownames(pvalues)
colnames(adj_pval) <- colnames(pvalues)
adj_pval <- adj_pval[order(rownames(adj_pval)),]

write.csv(data.frame(adj_pval), file = file.path(params$outputDir, paste0("09_ddbb_mcpas_", chain, ".csv")), row.names = T)
```


## ref db = vdjdb
```{r}
# db : reference database (vdjdb)
# cdr3 : data (CDR3 sets)
# disease : disease name (e.g. HIV, CMV, etc)
get_stats <- function(db, cdr3, disease) {
  disease_cdr3 <- unique(db$CDR3[db$Epitope.species == disease])
  cont_table <- table(unique(db$CDR3) %in% cdr3, unique(db$CDR3) %in% disease_cdr3)
  rownames(cont_table) <- c("not_in_data", "in_data")
  colnames(cont_table) <- c("not_in_disease","in_disease")
  pval <- fisher.test(cont_table, alternative = "great")$p.value
  return(pval)
}

pvalues = do.call(cbind, lapply(cdrsets, function(cdr3set){
  diseases <- unique(vdjdb_specie$Epitope.species)
  pvals <- rep(NA, length(diseases))

  for (i in 1:length(diseases)) {
    pvals[i] <- get_stats(db = vdjdb_specie, cdr3 = cdr3set, disease = diseases[i])
    names(pvals)[i] <- diseases[i]
  }

  return(pvals)

}))

adj_pval <- matrix(p.adjust(pvalues, method = "fdr"), nrow = nrow(pvalues), ncol = length(levels))
rownames(adj_pval) <- rownames(pvalues)
colnames(adj_pval) <- colnames(pvalues)
adj_pval = adj_pval[order(rownames(adj_pval)),]

write.csv(data.frame(adj_pval), file = file.path(params$outputDir, paste0("09_ddbb_vdjdb_", chain, ".csv")), row.names = T)
```
