---
title: "04 - Correlations"
author: "Teresa Rubio"
date: "March, 2021"
output:
    html_document:
      code_folding: hide
params:
  inputDir: !r getwd()
  workDir: !r getwd()
  outputDir: !r getwd()
  sampleInfo: !r NULL
  chain: !r NULL
---

# Packages
```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(ggpubr)
require(gridExtra)
library(grid)
```

# Work enviroment
```{r}
# Define output directory and TCR chain
dir.create(params$outputDir, recursive=T)
chain <- params$chain

# Load dataset overview info
DatasetInfo <- read.table("03_datasetoverview.csv", sep = ",", header = T, check.names = F)
levels <- unique(DatasetInfo$Group)
DatasetInfo$Group <- factor(DatasetInfo$Group, levels = levels)

# Color palette for sample groups
cbPalette <- c("#56B4E9","#CB181D","#000000","#CC79A7","#db6d00","#009E73",
               "#006ddb","#b66dff","#924900","#490092","#999999","#F0E442")
```


# Correlations
To examine whether the sequencing depth allowed adequate clonal coverage, we determined the correlation of the number of reads, the number of clones and SE vs. the number of cells and then between reads and clones number, reads number and SE, and clones and SE. Ideally, we expect the number of clones to be independent from the number of reads.

## Number of cells vs amount of RNA
A previous step as quality control is calculate the correlation between the total number of isolated CD4 cells vs. the total RNA amount extracted. Ideally, a high correlation is expected to be able to send enough RNA for sequencing.
```{r, message=FALSE, warning=FALSE}
# Scatter plot with correlation coefficient
p1 <- ggscatter(DatasetInfo, 
                x = "CellNumber", 
                y = "totalRNA",
                title="Pearson correlation",
                label = "SampleID",
                font.label = c(10,"black"),
                size = 3,
                add = "reg.line",
                color = "Group", 
                palette = cbPalette,
                conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3)  

# plots
p1
```

## Read vs. cellnumber
```{r, message=FALSE, warning=FALSE}
readsVScells <- ggscatter(DatasetInfo, 
                          x = paste0(chain,"reads"), 
                          y = "CellNumber",
                          label = "SampleID",
                          font.label = c(10,"black"),
                          size = 3,
                          add = "reg.line",
                          color = "Group", 
                          palette = cbPalette,
                          conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3)+
      theme(legend.position = 'none')

readsVScells
```

### Plot
```{r, results='hide', message=FALSE, warning=FALSE}
save(readsVScells, file = file.path(params$workDir, paste0("04_correlations_readsVScells_",chain,".Rda")))

png(file.path(params$outputDir, paste0("04_correlations_readsVScells_", chain, ".png")), 
    height = 30, width = 30, units = "cm", res = 300)
readsVScells
dev.off()
```

## Clones vs. cellnumber
```{r, message=FALSE, warning=FALSE}
clonesVScells <- ggscatter(DatasetInfo, 
                           x = paste0(chain,"clones"), 
                           y = "CellNumber",
                           label = "SampleID",
                           font.label = c(10,"black"),
                           size = 3,
                           add = "reg.line",
                           color = "Group", 
                           palette = cbPalette,
                           conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3) +
      theme(legend.position = 'none') 

clonesVScells
```

### Plot
```{r, results='hide', message=FALSE, warning=FALSE}
save(clonesVScells, file = file.path(params$workDir, paste0("04_correlations_clonesVScells_",chain,".Rda")))

png(file.path(params$outputDir, paste0("04_correlations_clonesVScells_", chain, ".png")),
    height = 30, width = 30, units = "cm", res = 300)
clonesVScells
dev.off()
```

## Clones vs. reads
```{r, message=FALSE, warning=FALSE}
clonesVSreads <- ggscatter(DatasetInfo, 
                           x = paste0(chain,"clones"), 
                           y = paste0(chain,"reads"),
                           label = "SampleID",
                           font.label = c(10,"black"),
                           size = 3,
                           add = "reg.line",
                           color = "Group", 
                           palette = cbPalette,
                           conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3) +
      theme(legend.position = 'right')

clonesVSreads
```

### Plot
```{r, results='hide', message=FALSE, warning=FALSE}
save(clonesVSreads, file = file.path(params$workDir, paste0("04_correlations_clonesVSreads_",chain,".Rda")))

png(file.path(params$outputDir, paste0("04_correlations_clonesVSreads_", chain, ".png")),
    height = 30, width = 30, units = "cm", res = 300)
clonesVSreads
dev.off()
```

## SE vs. cellnumber
```{r, message=FALSE, warning=FALSE}
seVScells <- ggscatter(DatasetInfo, 
                       x = "CellNumber", 
                       y =  paste0(chain,"shannonevenness"),
                       label = "SampleID",
                       font.label = c(10,"black"),
                       size = 3,
                       add = "reg.line",
                       color = "Group", 
                       palette = cbPalette,
                       conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3) +
      theme(legend.position = 'none')

seVScells
```

### Plot
```{r, results='hide', message=FALSE, warning=FALSE}
save(seVScells, file = file.path(params$workDir, paste0("04_correlations_seVScells_",chain,".Rda")))

png(file.path(params$outputDir, paste0("04_correlations_seVScells_", chain, ".png")),
    height = 30, width = 30, units = "cm", res = 300)
seVScells
dev.off()
```

## SE vs. clones
```{r, message=FALSE, warning=FALSE}
seVSclones <- ggscatter(DatasetInfo, 
                        x =  paste0(chain,"clones"), 
                        y =  paste0(chain,"shannonevenness"),
                        label = "SampleID",
                        font.label = c(10,"black"),
                        size = 3,
                        add = "reg.line",
                        color = "Group", 
                        palette = cbPalette,
                        conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3) +
      theme(legend.position = 'none') 

seVSclones
```

### Plot
```{r, results='hide', message=FALSE, warning=FALSE}
save(seVSclones, file = file.path(params$workDir, paste0("04_correlations_seVSclones_",chain,".Rda")))

png(file.path(params$outputDir, paste0("04_correlations_seVSclones_", chain, ".png")),
    height = 30, width = 30, units = "cm", res = 300)
seVSclones
dev.off()
```

## SE vs. reads
```{r, message=FALSE, warning=FALSE}
seVSreads <- ggscatter(DatasetInfo, 
                       x =  paste0(chain,"reads"), 
                       y =  paste0(chain,"shannonevenness"),
                       label = "SampleID",
                       font.label = c(10,"black"),
                       size = 3,
                       add = "reg.line",
                       color = "Group", 
                       palette = cbPalette,
                       conf.int = FALSE) +
      stat_cor(method = "pearson", aes(color = Group), label.x = 3)  +
      theme(legend.position = 'none')

seVSreads
```

### Plot
```{r, results='hide', message=FALSE, warning=FALSE}
save(seVSreads, file = file.path(params$workDir, paste0("04_correlations_seVSreads_",chain,".Rda")))

png(file.path(params$outputDir, paste0("04_correlations_seVSreads_", chain, ".png")),
    height = 30, width = 30, units = "cm", res = 300)
seVSreads
dev.off()
```


## Final plot
This is the final plot to include in the report:
```{r, results = 'hide', warning=FALSE, message=FALSE}
# Merged plot
png(file.path(params$outputDir, paste0("04_correlations_", chain, ".png")), 
    height = 30, width = 50, units = "cm", res = 300)
ggpubr::ggarrange(readsVScells, 
                  clonesVScells, 
                  clonesVSreads,
                  seVScells, 
                  seVSclones, 
                  seVSreads,
                  labels = c("A", "B", "C"),
                  ncol = 3, nrow = 3)
grid.arrange(readsVScells, clonesVScells, clonesVSreads,
             seVScells, seVSclones, seVSreads,
             ncol=3)
dev.off()
```
